### 2016-12-09 greedy_hamiltonian_revised/chapters/1.introduction.tex
Jan's comments included

\section{Introduction}
Parameterized partial differential equations are very often used as tools to model many problems in engineering and the applied sciences. While the need for more accuracy has led to the development of exceedingly complex models, the computational and storage limitations often make direct approaches impossible. Hence, we must seek alternative methods that allow us to approximate the desired output under variation of the input parameters while keeping the computational costs to a minimum.

Reduced basis methods have emerged as a powerful approach for the reduction of the intrinsic complexity of such models \cite{Ito:1998up,Ito:1998ch,Ito:2001ev,Peterson:1989ki}. These methods contain two stages: the offline and the online. In the offline stage, one explors the parameter space to construct a low-dimensional basis that accurately represents the parametrized solution to the partial differential equation. In this stage, evaluation of the solution of the original model for multiple parameter values is required. The online stage comprises a Galerkin projection onto the span of the reduced basis, which allows exploration of the parameter space at significantly reduced complexity \cite{Antoulas:2005:ALD:1088857,Anonymous:2016wl}.

Convectional reduced basis techniques e.g. proper orthogonal decomposition (POD) \cite{Kunisch:2002er,Atwell:2001ja,Ravindran:2002hn}, require the exploration of the entire parameter space. This leads to a very expensive and often impractical offline stages when dealing with multi-dimensional paramtere domains. On the other hand, sampling techniques, usually of a greedy nature, search through the parameter space selectively often guided by an error estimate to certify the accuracy of the basis. This approach, accompanied with an efficient sampling procedure, trades considerable amount of computation with overall accuracy of the reduced-basis \cite{Cuong:2005gd,Rozza:2005ie,Anonymous:2016wl}.

Beside computational complexity, another aspect of reduced order modeling is the preservation of structures and, in particular, stability of the original model. In general, reduced order models do not guarantee that such properties are preserved \cite{Anonymous:pMn0O0Q4}.

In the context of Hamiltonian and Lagrangian systems, recent work suggests modifications of POD to preserve geometric structure. Lall et al. \cite{Lall:2003iy} and Carlberg et al. \cite{Carlberg:2014ky} suggests that the reduced-order system should be identified by a Lagrangian function on a low-ordered configuration space. In this way the geometric structure of the original system is inherited by the reduced system. For Hamiltonian systems Peng et al. \cite{Peng:2014di}, using a symplectic transformation, constructs a reduced Hamiltonian, as an approximation to the Hamiltonian of the original system. As a result the reduced system preserves the symplectic structure. Although these methods preserve geometric structure, they use a POD-like approach for constructing the reduced basis, and are not well suitable for problems with a high-dimensional parameter domain.

In this paper, we present a greedy approach for the construction of a reduced system that preserves the geometric structure of Hamiltonian systems, as an iterative routine that successively enriches the reduced basis at each iteration. On one hand this technique results in a reduced Hamiltonian system that mimics symplectic properties of the original system and preserves the Hamiltonian structure and stability over the course of time integration. On the other hand, since time integration of the original system is only required once per iteration, the proposed method saves substantial computational cost during the offline stage when compared to alternative POD-like approaches. Moreover, we demonstrate that assumptions, natural for the set of all solutions of the original Hamiltonian system under variation of parameters, lead to exponentially fast convergence of the greedy algorithm. For nonlinear Hamiltonian systems, we show how the basis can be combined with the discrete empirical interpolation method (DEIM) to allow a fast evaluation of nonlinear terms while maintaining the symplectic structure.

This paper is organized as follows. Section \ref{chap:MoOr:1} presents a brief overview of model order reduction, POD and DEIM. In Section \ref{chap:Hasy:1} we cover the required topics from symplectic geometry and Hamiltonian systems. Section \ref{chap:SyMo:1} discusses the greedy generation of a symplectic reduced basis as well as other SVD-based symplectic model reduction techniques. Accuracy, stability, and efficiency of the greedy method compared to other SVD-based methods are discussed in Section \ref{chap:NuRe:1}. Finally we offer the conclusion in Section \ref{chap:Con:1}.

>>>>

\section{Introduction}
{\edit Parameterized partial differential equations often arise as a model in many problems in engineering and the applied sciences}. While the need for more accuracy has led to the development of exceedingly complex models, the {\edit limitations in computational cost and storage} often make direct approaches {\edit impractical}. Hence, we must seek alternative methods that allow us to approximate the desired output under variation of the input parameters while keeping the computational costs to a minimum.

Reduced basis methods have emerged as a powerful approach for the reduction of the intrinsic complexity of such models \cite{Ito:1998up,Ito:1998ch,Ito:2001ev,Peterson:1989ki}. These methods contain two stages: {\edit the offline stage and the online stage}. In the offline stage, one explores the parameter space to construct a low-dimensional basis that accurately represents the parametrized solution to the partial differential equation. In this stage, the evaluation of the solution of the original model for multiple parameter values is required. The online stage comprises a Galerkin projection onto the span of the reduced basis, which allows exploration of the parameter space at a significantly reduced complexity \cite{Antoulas:2005:ALD:1088857,Anonymous:2016wl}.

Convectional reduced basis techniques, such as proper orthogonal decomposition (POD) \cite{Kunisch:2002er,Atwell:2001ja,Ravindran:2002hn}, require the exploration of the entire parameter space. This leads to a very expensive and often impractical offline {\edit stage} when dealing with multi-dimensional parameter domains. On the other hand, sampling techniques, usually of a greedy nature, search through the parameter space selectively, guided by an error estimate to certify the accuracy of the basis. This approach, accompanied with an efficient sampling procedure, {\edit balances the cost} of computation with the overall accuracy of the reduced-basis \cite{Cuong:2005gd,Rozza:2005ie,Anonymous:2016wl}.

{\edit Besides} computational complexity, another aspect of reduced order modeling is the preservation of structure and, in particular, {\edit the} stability of the original model. In general, reduced order models do not guarantee that such properties are preserved \cite{Anonymous:pMn0O0Q4}.

In the context of Hamiltonian and Lagrangian systems, recent work suggests modifications of POD to preserve {\edit some} geometric structure. Lall et al. \cite{Lall:2003iy} and Carlberg et al. \cite{Carlberg:2014ky} suggests that the reduced-order system should be identified by a Lagrangian function on a low-dimensional configuration space. In this way, the geometric structure of the original system is inherited by the reduced system. {\edit Model reduction for port-Hamiltonian systems can be found in the works of Beattie et al. \cite{Chaturantabut:2016he}, Polyuga et al. \cite{Polyuga:2010gj} and references therein. These works construct a reduced port-Hamiltonian system using Krylov or POD methods that inherit the passivity and stability of the original system.} For Hamiltonian systems, Peng et al. \cite{Peng:2014di}, using a symplectic transformation, constructs a reduced Hamiltonian, as an approximation to the Hamiltonian of the original system. As a result, the reduced system preserves the symplectic structure. Although these methods preserve {\edit the} geometric structure, they use a POD-like approach for constructing the reduced basis and are not well {\edit suited} for problems with a high-dimensional parameter domain.

In this paper, we present a greedy approach for the construction of a reduced system that preserves the geometric structure of Hamiltonian systems. This technique results in a reduced Hamiltonian system that mimics the symplectic properties of the original system and preserves the Hamiltonian structure and stability over the course of time. On the other hand, since time integration of the original system is only required once per iteration, the proposed method saves substantial computational cost during the offline stage when compared to alternative POD-like approaches. {\edit It is well known that structured matrices, e.g. symplectic matrices, generally are not well-conditioned \cite{Karow:2006cf}. The greedy update of the symplectic basis presented here, yields a orthosymplectic basis and, therefore, a norm bounded basis.} Moreover, we demonstrate that assumptions, natural for the set of all solutions of the original Hamiltonian system under variation of parameters, lead to exponentially fast convergence of the greedy algorithm. For nonlinear Hamiltonian systems, we show how the basis can be combined with the discrete empirical interpolation method (DEIM) {\edit \cite{Chaturantabut:2010cz,Barrault:2004kz}} to enable a fast evaluation of nonlinear terms while maintaining the symplectic structure.

This paper is organized as follows. Section \ref{chap:MoOr:1} presents a brief overview of model order reduction, POD and DEIM. In Section \ref{chap:Hasy:1} we cover the required topics from symplectic geometry and Hamiltonian systems. Section \ref{chap:SyMo:1} discusses the greedy generation of a symplectic reduced basis as well as other SVD-based symplectic model reduction techniques. Accuracy, stability, and efficiency of the greedy method compared to other SVD-based methods are discussed in Section \ref{chap:NuRe:1}. {\edit Finally we offer some conclusive remarks in Section \ref{chap:Con:1}}.

### 2016-12-09 greedy_hamiltonian_revised/chapters/2.mor.tex
all done, response to the referees also added
Jan's comments included

\section{Model Order Reduction} \label{chap:MoOr:1}
Consider a parameterized, finite dimensional dynamical system that can be described by a set of first order ordinary differential equations on the form
\begin{equation} \label{eq:MoOr:1}
\left\{
\begin{split}
& \frac{d}{dt} \mathbf{x}(t,\omega) = \mathbf f (t,\mathbf x,\omega), \\
& \mathbf x(0,\omega) = \mathbf x_0(\omega).
\end{split}
\right.
\end{equation}
Here $\mathbf x \in \mathbb R^n$ is the state vector, $\omega \in \Gamma$ is a vector containing all the parameters of the system that belongs to a compact set $\Gamma$ ($\subset \mathbb R^d$) and $\mathbf f : \mathbb R \times \mathbb R^n \times \Gamma \to \mathbb R^n$ is a general function of the state variables and parameters.

We define the solution manifold as the set of all solutions to (\ref{eq:MoOr:1}) under variation of {\edit the parameters in $\Gamma$}
\begin{equation} \label{eq:MoOr:2}
	\mathcal M = \{ \mathbf x(\omega) | \omega \in \Gamma \} \subset \mathbb R^n.
\end{equation}
Note that the exact solution and solution manifold is not always available; we assume that we have a numerical integrator in hand that can approximate the solution to (\ref{eq:MoOr:1}) for any realization of $\omega$ {\edit with a given accuracy}. By abuse of notation we refer to $\mathbf x$ and $\mathcal M$ as the exact solution and the exact solution manifold, respectively, rather than the discrete solution and discrete solution manifold.

Model order reduction is based on the assumption that $\mathcal M$ is of low dimension \cite{Anonymous:2016wl,Antoulas:2005:ALD:1088857}, and that the span of appropriately chosen basis vectors $\{v_i\}_{i=1}^k$ covers most of the solution manifold with a small error. The set $\{v_i\}_{i=1}^k$ is denoted as the \emph{reduced basis} and its span as the \emph{reduced space}. Assuming that a $k$-dimensional $(k\ll n)$ reduced basis is given, the approximated solution can be represented as
\begin{equation} \label{eq:MoOr:3}
	\mathbf x \approx V \mathbf y,
\end{equation}
where $V$ is a matrix containing the reduced basis vectors as its columns and $\mathbf y$ is the coordinates of the approximation in this basis. By substituting (\ref{eq:MoOr:3}) into (\ref{eq:MoOr:1}) we obtain the overdetermined system
\begin{equation} \label{eq:MoOr:4}
	V \frac{d}{dt} \mathbf y = \mathbf f (t , V \mathbf y , \omega) + \mathbf r(t,\omega).
\end{equation}
Here we added the residual $\mathbf r$ to emphasise that (\ref{eq:MoOr:4}) is an approximation of (\ref{eq:MoOr:1}). Taking the Petrov-Galerkin projection \cite{Antoulas:2005:ALD:1088857} we construct a basis $W$ of size $n-k$ that is orthogonal to the residual $\mathbf r$ and assume $W^T V$ is invertible. This yields
\begin{equation} \label{eq:MoOr:5}
	\frac{d}{dt} \mathbf y = (W^TV)^{-1} \mathbf f(t,V\mathbf y,\omega).
\end{equation}
Equation (\ref{eq:MoOr:5}) consists of $k$ equations and is called the reduced system. Solving the reduced system instead of the original system can reduce the computational costs since $k$ is significantly smaller than $n$. For nonlinear systems, the evaluation of $\mathbf f$ may still have computational complexity that depends on $n$. We return to this question in detail in Section \ref{chap:MoOr.DEIM:1}.

\subsection{Proper Orthogonal Decomposition} \label{chap:MoOr.PrOr:1}
Let $\mathbf x (t_i,\omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$ be a finite number of samples, referred to as \emph{snapshots}, from the solution manifold (\ref{eq:MoOr:2}). If we {\edit suppose} that a reduced basis $V$ is provided, the projection operator from $\mathbb R^n$ onto the reduced space can be constructed as $VV^T$. The proper orthogonal decomposition (POD) requires the total error of projecting all the snapshots onto the reduced space to be minimum. The POD basis of size $k$ is thus the solution of the optimization problem
\begin{equation} \label{eq:MoOr:6}
\begin{aligned}
& \underset{V\in \mathbb R^{n\times k}}{\text{minimize}}
& & \| S - VV^TS\|_F \\
& \text{subject to}
& & V^TV = I_k
\end{aligned}
\end{equation}
Here $S$ is the snapshot matrix, containing snapshots $\mathbf x(t_i,\omega_j)$ in its columns, $\|\cdot \|_F$ is the Frobenius norm of a matrix and $I_k$ is the identity matrix of size $k$. According to {\edit Schmidt-Mirsky-Eckart-Young theorem \cite{Markovsky:2011:LRA:2103589}}, the solution to the optimization (\ref{eq:MoOr:6}) is equivalent to the truncated singular value decomposition (SVD) of the snapshot matrix $S$ given by
\begin{equation} \label{eq:MoOr:7}
	V = \sigma_1 u_1 v^T_1 + \cdots + \sigma_k u_k v^T_k.
\end{equation}
Here $\sigma_i, u_i$ and $v_i$ are the singular values, left singular vectors and right singular vectors of $S$, respectively {\edit \cite{Markovsky:2011:LRA:2103589} }.

\subsection{Discrete Empirical Interpolation Method (DEIM)} \label{chap:MoOr.DEIM:1} \nocite{Chaturantabut:2010cz}
In this section we {\edit discuss the efficiency} of evaluating nonlinearities when using projection based reduced model. Suppose that the right hand side function in (\ref{eq:MoOr:1}) is of the form $\mathbf f(t,\mathbf x , \omega) = L\mathbf x + \mathbf g(t,\mathbf x ,\omega)$, where $L\in \mathbb R^{n\times n}$ reflects the linear part, and $\mathbf g$ is a nonlinear function. Now suppose a $k$-dimensional reduced basis $V$ is provided. The reduced system takes the form
\begin{equation} \label{eq:MoOr:8}
	\frac{d}{dt} \mathbf y = \underbrace{(WV)^{-1} L V}_{\tilde L} \mathbf{y} + \underbrace{(WV)^{-1} \mathbf g(t,V\mathbf y,\omega)}_{\tilde N (\mathbf y)}.
\end{equation}
Here, $\tilde L$ is a $k\times k$ matrix which can be computed before time integration of the reduced system. However, evaluation of $\tilde N (\mathbf y)$ has a complexity that depends on $n$, the size of the original system. Suppose that evaluation of $\mathbf g$ with $n$ components has complexity $\alpha(n)$, for some function $\alpha$. Then the complexity of evaluating $\tilde N(\mathbf y)$ is $\mathcal{O}(\alpha(n) + 4nk)$ which consists of 2 matrix-vector operations and the evaluation of the nonlinear function. This means that estimation of the nonlinear terms can be as expensive as solving the original system.

{\edit To overcome this bottleneck we take an approach similar to section \ref{chap:MoOr.PrOr:1} \cite{Chaturantabut:2010cz,Barrault:2004kz}}. Assume that the manifold $\mathcal M_{\mathbf g} = \{ \mathbf g(t,\mathbf x , \omega)| t\in \mathbb R, \mathbf x \in \mathbb R , \omega \in \Gamma\}$ is of a low dimension and that $\mathbf g$ can be approximated by a linear subspace of dimension $m\ll n$, spanned by the basis $\{ u_1 , \dots , u_m \}$, i.e.
\begin{equation} \label{eq:MoOr:10}
	\mathbf g(t,\mathbf x,\omega) \approx U \mathbf c(t,\mathbf x,\omega).
\end{equation}
Here $U$ contains basis vectors $u_i$ and $\mathbf c$ is the corresponding vector of coefficients. Now suppose $p_1,\dots,p_m$ are $m$ indices from $\{1,\dots,n\}$ and define an $n\times m$ matrix
\begin{equation} \label{eq:MoOr:11}
	P = [e_{p_1},\dots,e_{p_m}],
\end{equation}
where $e_{p_i}$ is the $p_i$-th column of the identity matrix $I_n$. Hence, multiplying $P$ with $\mathbf g$ selects components $p_1,\dots,p_m$ of $\mathbf g$. If we assume that $P^TU$ is non-singular, the coefficient vector $\mathbf c$ can be uniquely determined from
\begin{equation} \label{eq:MoOr:12}
	P^T \mathbf g = (P^TU)\mathbf c.
\end{equation}
Finally the approximation of $\mathbf g$ is determined by
\begin{equation} \label{eq:MoOr:13}
	\mathbf g(t,\mathbf x,\omega) \approx U \mathbf c(t,\mathbf x,\omega) = U (P^TU)^{-1} P^T \mathbf g(t,\mathbf x,\omega),
\end{equation}
which is referred to as the \emph{Discrete Empirical Interpolation} (DEIM) approximation \cite{Chaturantabut:2010cz}. Applying DEIM on the reduced system (\ref{eq:MoOr:5}) yields
\begin{equation} \label{eq:MoOr:14}
	\frac{d}{dt} \mathbf y = \tilde L \mathbf y + (WV)^{-1} U(P^TU)^{-1}P^T \mathbf g(t,V\mathbf y , \omega).
\end{equation}
Note that the matrix $(WV)^{-1} U(P^TU)^{-1}$ can be computed offline and since $\mathbf g$ is evaluated only at $m$ of its components, the evaluation of the nonlinear term in (\ref{eq:MoOr:14}) does not depend on $n$.

{\edit To obtain the projection basis $U$, the POD can be applied to the \emph{nonlinear snapshots}. The nonlinear snapshots are the ensemble of samples of the nonlinear term $\mathbf g(t_i,\mathbf x, \omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$}. There is no additional cost for computing the nonlinear snapshots, since they are generated when computing the trajectory snapshot matrix $S$. The interpolating indices $p_1,\dots,p_m$ can be constructed as follows. Given the projection basis $U = \{u_1,\dots,u_m\}$, the first interpolation index $p_1$ is chosen according to the component of $u_1$ with the largest magnitude. The rest of the interpolation indices, $p_2,\dots,p_m$ correspond to the component of the largest magnitude of the residual vector $\mathbf r = u_l - U \mathbf c$. It is shown in \cite{Chaturantabut:2010cz} that if the residual vector is a nonzero vector in each iteration then $P^TU$ is non-singular and (\ref{eq:MoOr:13}) is well defined. We summarize the algorithm of generating interpolating indices in Algorithm \ref{alg:MoOr:1}.

@@

{\edit Numerical solution of (\ref{eq:MoOr:8}) sometimes involve computation of the Jacobian of the nonlinear function $\mathbf g(t,\mathbf x, \omega)$ with respect to the reduced state variable $\mathbf y$}
\begin{equation} \label{eq:MoOr:9}
	\edit \mathbf J_{\mathbf x}(\mathbf g) = (WV)^{-1} \mathbf J_{\mathbf y}(\mathbf g) V,
\end{equation}
{\edit where $\mathbf J_\alpha(\mathbf g)$ is the Jacobian matrix of $\mathbf g$ with respect to the variable $\alpha$.} The complexity of computing (\ref{eq:MoOr:9}) is $\mathcal{O}(\alpha(n) +2n^2k+2nk^2+2nk)$ comprising several matrix-vector multiplications and an evaluation of the Jacobian which depends on the size of the original system. Approximating the Jacobian in (\ref{eq:MoOr:9}) is usually both problem and discretization dependent. Very often the nonlinear function $\mathbf g$ is evaluated component-wise i.e.
\begin{equation} \label{eq:MoOr:15}
	\mathbf g(\mathbf x) =
	\begin{pmatrix}
		g_1(x_1,\dots,x_n) \\
		g_2(x_1,\dots,x_n) \\
		\vdots \\
		g_n(x_1,\dots,x_n)
	\end{pmatrix}
	=
	\begin{pmatrix}
		g_1(x_1) \\
		g_2(x_2) \\
		\vdots \\
		g_n(x_n)
	\end{pmatrix}.
\end{equation}
with $\mathbf x$ the configuration vector. In such cases the interpolating index matrix $P$ and the nonlinear function $\mathbf g$ commute, i.e.,
\begin{equation} \label{eq:MoOr:16}
	\tilde N(\mathbf y) \approx (WV)^{-1} U(P^TU)^{-1}P^T \mathbf g(V\mathbf y) = (WV)^{-1} U(P^TU)^{-1}\mathbf g(P^TV\mathbf y)
\end{equation}
If we now take the Jacobian of the approximate function we recover
\begin{equation}
	\edit \mathbf J_{\mathbf x}(\mathbf g) = \underbrace{ (WV)^{-1} U(P^TU)^{-1} }_{k\times m} \underbrace{ \mathbf J_{\mathbf y}(\mathbf g(P^T V \mathbf y) ) }_{m\times m} \underbrace{P^T V}_{m\times k}.
\end{equation}
The matrix $(WV)^{-1} U(P^TU)^{-1}$ can be computed offline and the Jacobian is evaluated only at $m\times m$ components. Hence the overall complexity of computing the Jacobian is now independent of $n$. {\edit We refer the reader to \cite{Barrault:2004kz,Chaturantabut:2010cz} for more detail.}

>>>>

\section{Model Order Reduction} \label{chap:MoOr:1}
Consider a parameterized, finite dimensional dynamical system described by a set of first order ordinary differential equations
\begin{equation} \label{eq:MoOr:1}
\left\{
\begin{split}
& \frac{d}{dt} \mathbf{x}(t,\omega) = \mathbf f (t,\mathbf x,\omega), \\
& \mathbf x(0,\omega) = \mathbf x_0(\omega).
\end{split}
\right.
\end{equation}
Here $\mathbf x \in \mathbb R^n$ is the state vector, $\omega \in \Gamma$ is a vector containing all the parameters of the system {\edit belonging} to a compact set $\Gamma$ ($\subset \mathbb R^d$) and $\mathbf f : \mathbb R \times \mathbb R^n \times \Gamma \to \mathbb R^n$ is a general {\edit vector valued} function of the state variables and parameters.

We define the solution manifold as the set of all solutions to (\ref{eq:MoOr:1}) under variation of {\edit the parameters in $\Gamma$}
\begin{equation} \label{eq:MoOr:2}
	\edit \mathcal M = \{ \mathbf x(t,\omega) | \omega \in \Gamma,\ t\geq 0 \} \subset \mathbb R^n.
\end{equation}
Note that the exact solution and solution manifold is {\edit often not} available; we assume that we have a numerical integrator that can approximate the solution to (\ref{eq:MoOr:1}) for any realization of $\omega$ {\edit with a given accuracy}. By abuse of notation, we refer to $\mathbf x$ and $\mathcal M$ as the exact solution and the exact solution manifold, respectively, rather than the discrete solution and discrete solution manifold.

Model order reduction is based on the assumption that $\mathcal M$ is of low dimension \cite{Anonymous:2016wl,Antoulas:2005:ALD:1088857} and that the span of appropriately chosen basis vectors $\{v_i\}_{i=1}^k$ covers most of the solution manifold {\edit to within} a small error. The set $\{v_i\}_{i=1}^k$ is denoted as the \emph{reduced basis} and its span as the \emph{reduced space}. Assuming that a $k$-dimensional $(k\ll n)$ reduced basis is given, the approximated solution can be represented as
\begin{equation} \label{eq:MoOr:3}
	\mathbf x \approx V \mathbf y,
\end{equation}
where $V$ is a matrix containing the reduced basis vectors as its columns and $\mathbf y$ {\edit contains} the coordinates of the approximation in this basis. By substituting (\ref{eq:MoOr:3}) into (\ref{eq:MoOr:1}) we obtain the overdetermined system
\begin{equation} \label{eq:MoOr:4}
	V \frac{d}{dt} \mathbf y = \mathbf f (t , V \mathbf y , \omega) + \mathbf r(t,\omega).
\end{equation}
Here we added the residual $\mathbf r$ to emphasize that (\ref{eq:MoOr:4}) is an approximation of (\ref{eq:MoOr:1}). Taking the Petrov-Galerkin projection \cite{Antoulas:2005:ALD:1088857} we construct a basis $W$ of size $n-k$ that is orthogonal to the residual $\mathbf r$ and {\edit requires that} $W^T V$ is invertible. This yields
\begin{equation} \label{eq:MoOr:5}
	\frac{d}{dt} \mathbf y = (W^TV)^{-1} \mathbf f(t,V\mathbf y,\omega).
\end{equation}
Equation (\ref{eq:MoOr:5}) consists of $k$ equations and is called the reduced system. Solving the reduced system instead of the original system can reduce the computational costs {\edit provided} $k$ is significantly smaller than $n$. For nonlinear systems, the evaluation of $\mathbf f$ may still have computational complexity that depends on $n$. We return to this question in detail in Section \ref{chap:MoOr.DEIM:1}.

\subsection{Proper Orthogonal Decomposition} \label{chap:MoOr.PrOr:1}
Let $\mathbf x (t_i,\omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$ be a finite number of samples, referred to as \emph{snapshots}, from the solution manifold (\ref{eq:MoOr:2}). If we {\edit assume} that a reduced basis $V$ is provided, the projection operator from $\mathbb R^n$ onto the reduced space can be constructed as $VV^T$. The proper orthogonal decomposition (POD) requires the total error of projecting all the snapshots onto the reduced space to be {\edit minimized}. The POD basis of size $k$ is thus the solution to the optimization problem
\begin{equation} \label{eq:MoOr:6}
\begin{aligned}
& \underset{V\in \mathbb R^{n\times k}}{\text{minimize}}
& & \| S - VV^TS\|_F \\
& \text{subject to}
& & V^TV = I_k
\end{aligned}
\end{equation}
Here $S$ is the snapshot matrix, containing snapshots $\mathbf x(t_i,\omega_j)$ in its columns, $\|\cdot \|_F$ is the Frobenius norm and $I_k$ is the identity matrix of size $k$. According to {\edit Schmidt-Mirsky-Eckart-Young theorem \cite{Markovsky:2011:LRA:2103589}}, the solution to (\ref{eq:MoOr:6}) is equivalent to the truncated singular value decomposition (SVD) of the snapshot matrix $S$ given by
\begin{equation} \label{eq:MoOr:7}
	V = \sigma_1 u_1 v^T_1 + \cdots + \sigma_k u_k v^T_k.
\end{equation}
Here $\sigma_i, u_i$ and $v_i$ are the singular values, the left singular vectors, and the right singular vectors of $S$, respectively {\edit \cite{Markovsky:2011:LRA:2103589} }.

\subsection{Discrete Empirical Interpolation Method (DEIM)} \label{chap:MoOr.DEIM:1} \nocite{Chaturantabut:2010cz}
In this section we {\edit discuss the efficiency} of evaluating nonlinearities {\edit in the context of} projection based reduced models. Suppose that the right hand side in (\ref{eq:MoOr:1}) is of the form $\mathbf f(t,\mathbf x , \omega) = L\mathbf x + \mathbf g(t,\mathbf x ,\omega)$, where $L\in \mathbb R^{n\times n}$ reflects the linear part, and $\mathbf g$ is a nonlinear function. Now {\edit assume that} a $k$-dimensional reduced basis $V$ is provided. The reduced system takes the form
\begin{equation} \label{eq:MoOr:8}
	\frac{d}{dt} \mathbf y = \underbrace{(WV)^{-1} L V}_{\tilde L} \mathbf{y} + \underbrace{(WV)^{-1} \mathbf g(t,V\mathbf y,\omega)}_{\tilde N (\mathbf y)}.
\end{equation}
Here, $\tilde L$ is a $k\times k$ matrix which can be computed before time integration of the reduced system. However, {\edit the} evaluation of $\tilde N (\mathbf y)$ has a complexity that depends on $n$, the size of the original system. Suppose that the evaluation of $\mathbf g$ with $n$ components has the complexity $\alpha(n)$, for some function $\alpha$. Then the complexity of evaluating $\tilde N(\mathbf y)$ is $\mathcal{O}(\alpha(n) + 4nk)$ which consists of 2 matrix-vector operations and the {\edit evaluation of the nonlinear function, i.e. the evaluation of the nonlinear terms can be as expensive as solving the original system.}

{\edit To overcome this bottleneck we take an approach similar to that of Section \ref{chap:MoOr.PrOr:1} \cite{Chaturantabut:2010cz,Barrault:2004kz}}. Assume that the manifold $\mathcal M_{\mathbf g} = \{ \mathbf g(t,\mathbf x , \omega)| t\in \mathbb R, \mathbf x \in \mathbb R , \omega \in \Gamma\}$ is of a low dimension and that $\mathbf g$ can be approximated by a linear subspace of dimension $m\ll n$, spanned by the basis $\{ u_1 , \dots , u_m \}$, i.e.
\begin{equation} \label{eq:MoOr:10}
	\mathbf g(t,\mathbf x,\omega) \approx U \mathbf c(t,\mathbf x,\omega).
\end{equation}
Here $U$ contains basis vectors $u_i$ and $\mathbf c$ is the vector of coefficients. Now suppose $p_1,\dots,p_m$ are $m$ indices from $\{1,\dots,n\}$ and define an $n\times m$ matrix
\begin{equation} \label{eq:MoOr:11}
	P = [e_{p_1},\dots,e_{p_m}],
\end{equation}
where $e_{p_i}$ is the $p_i$-th column of the identity matrix $I_n$. Multiplying $P$ with $\mathbf g$ selects components $p_1,\dots,p_m$ of $\mathbf g$. If we assume that $P^TU$ is non-singular, the coefficient vector $\mathbf c$ can be uniquely determined from
\begin{equation} \label{eq:MoOr:12}
	P^T \mathbf g = (P^TU)\mathbf c.
\end{equation}
Finally the approximation of $\mathbf g$ is determined by
\begin{equation} \label{eq:MoOr:13}
	\mathbf g(t,\mathbf x,\omega) \approx U \mathbf c(t,\mathbf x,\omega) = U (P^TU)^{-1} P^T \mathbf g(t,\mathbf x,\omega),
\end{equation}
which is referred to as the \emph{Discrete Empirical Interpolation} (DEIM) approximation \cite{Chaturantabut:2010cz}. Applying DEIM to the reduced system (\ref{eq:MoOr:5}) yields
\begin{equation} \label{eq:MoOr:14}
	\frac{d}{dt} \mathbf y = \tilde L \mathbf y + (WV)^{-1} U(P^TU)^{-1}P^T \mathbf g(t,V\mathbf y , \omega).
\end{equation}
Note that the matrix $(WV)^{-1} U(P^TU)^{-1}$ can be computed offline and since $\mathbf g$ is evaluated only at $m$ of its components, the evaluation of the nonlinear term in (\ref{eq:MoOr:14}) does not depend on $n$.

{\edit To obtain the projection basis $U$, the POD can be applied to the ensemble of samples of the nonlinear term $\mathbf g(t_i,\mathbf x, \omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$}. There is no additional cost {\edit associated with} computing the nonlinear snapshots, since they are generated when computing the trajectory snapshot matrix $S$. The interpolating indices $p_1,\dots,p_m$ can be constructed as follows. Given the projection basis $U = \{u_1,\dots,u_m\}$, the first interpolation index $p_1$ is chosen according to the component of $u_1$ with the largest magnitude. The rest of the interpolation indices, $p_2,\dots,p_m$ correspond to the component of the largest magnitude of the residual vector $\mathbf r = u_l - U \mathbf c$. It is shown in \cite{Chaturantabut:2010cz} that if the residual vector is a nonzero vector in each iteration then $P^TU$ is non-singular and (\ref{eq:MoOr:13}) is well defined.

@@

{\edit The numerical solution of (\ref{eq:MoOr:8}) may involve the computation of the Jacobian of the nonlinear function $\mathbf g(t,\mathbf x, \omega)$ with respect to the reduced state variable $\mathbf y$}
\begin{equation} \label{eq:MoOr:9}
	\edit \mathbf J_{\mathbf y}(\mathbf g) = (WV)^{-1} \mathbf J_{\mathbf x}(\mathbf g) V,
\end{equation}
{\edit where $\mathbf J_\alpha(\mathbf g)$ is the Jacobian matrix of $\mathbf g$ with respect to the variable $\alpha$.} The complexity of (\ref{eq:MoOr:9}) is $\mathcal{O}(\alpha(n) +2n^2k+2nk^2+2nk)$, comprising several matrix-vector multiplications and an evaluation of the Jacobian which depends on the size of the original system. Approximating the Jacobian in (\ref{eq:MoOr:9}) is usually both problem and discretization dependent. Often the nonlinear function $\mathbf g$ is evaluated component-wise i.e.
\begin{equation} \label{eq:MoOr:15}
	\mathbf g(\mathbf x) =
	\begin{pmatrix}
		g_1(x_1,\dots,x_n) \\
		g_2(x_1,\dots,x_n) \\
		\vdots \\
		g_n(x_1,\dots,x_n)
	\end{pmatrix}
	=
	\begin{pmatrix}
		g_1(x_1) \\
		g_2(x_2) \\
		\vdots \\
		g_n(x_n)
	\end{pmatrix}.
\end{equation}
In such cases the interpolating index matrix $P$ and the nonlinear function $\mathbf g$ commute, i.e.,
\begin{equation} \label{eq:MoOr:16}
	\tilde N(\mathbf y) \approx (WV)^{-1} U(P^TU)^{-1}P^T \mathbf g(V\mathbf y) = (WV)^{-1} U(P^TU)^{-1}\mathbf g(P^TV\mathbf y)
\end{equation}
If we now take the Jacobian of the approximate function we recover
\begin{equation}
	\edit \mathbf J_{\mathbf y}(\mathbf g) = \underbrace{ (WV)^{-1} U(P^TU)^{-1} }_{k\times m} \underbrace{ \mathbf J_{\mathbf x}(\mathbf g(P^T V \mathbf y) ) }_{m\times m} \underbrace{P^T V}_{m\times k}.
\end{equation}
The matrix $(WV)^{-1} U(P^TU)^{-1}$ can be computed offline and the Jacobian is evaluated only for $m\times m$ components. Hence the overall complexity of computing the Jacobian is now independent of $n$. {\edit We refer the reader to \cite{Barrault:2004kz,Chaturantabut:2010cz} for more detail.}
